Probability Aggregation: Dynamic Hierarchical Modeling of Expert Beliefs in Geopolitical Events

### 2013-08-24 paper.tex
Edits

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. This means that the decision-maker faces a different group of experts  every day, with only a few experts returning later on for a second round of advice. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point. As a result, standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not directly applicable to our setting. In response to this lack of appropriate procedures, we introduce an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is then used for
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

The paper begins with a description of our probability forecasting data that were collected by asking over 2,300 experts to give probability forecasts and to self-assess their level of expertise on a subset of 166 geopolitical binary events. After summarizing the dataset, the paper introduces a dynamic hierarchical model for capturing the crowd belief. The model is estimated in a two-step procedure. First, constrained parameter estimates are obtained via Gibbs sampling (see \citet{geman1984stochastic} for the original introduction of Gibbs sampling). These estimates are then transformed to their unconstrained equivalents by a calibration step. An extension of this model to polychotomous outcomes is briefly discussed before model evaluation. The first evaluation section uses synthetic data to study the accuracy to which the two-step procedure is able to estimate parameter values. The second evaluation section discusses different aspects of the model by applying it to our real-world probability forecasting data. The paper concludes with a discussion on future research directions and model limitations.

@@

\begin{figure}[h!]
\centering
\hspace*{0em} 	\includegraphics[width=  \textwidth]{Figures/LegendExamplePlot} % requires the graphicx package
\vspace{-4.5em}

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic time-series setting in which probability forecasts are made very infrequently by a heterogeneous group of experts. The resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response to this lack of appropriate methodology, our work 
introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated crowd belief across time. The model is estimated in two-steps: The first step, known as the \textit{sampling step}, samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions. The second step, known as the \textit{calibration step}, involves a one-dimensional optimization procedure that transforms the constrained parameter values into their unconstrained counterparts.

%Due to the unconventional nature of our dataset, our work expands the literature towards less-studied areas of  probability aggregation.

\subsection{Summary of Findings}
The model was applied to an unusually large dataset on probability forecasting done by human experts. The estimated crowd belief was found to be sharp and well-calibrated under both in- and out-of-sample probability estimation. This has direct implications  on predictive power and model interpretability. Firstly, the model was shown to outperform other probability aggregators in terms of forecasting ability. Secondly, the crowd belief was used as the no-bias reference point to study the distributions of the bias terms for different expertise-groups. All the groups were found to be under-confident. The under-confidence, however, decreased as the level of self-reported expertise increased. This result is about groups of experts and hence does not conflict with the well-known result of the individuals being over-confident (see, e.g.,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications}). Besides making predictions or studying group-level bias, the model can be used to generate estimates of many problem-specific quantities. These quantities have clear interpretations and can be combined in many interesting ways to answer a range of hypotheses about different types of questions and expert behavior.

\subsection{Future Directions and Limitations}
This paper developed a time-series model that is parsimonious subject to addressing the main challenges that arise from sparse probability forecasting data. The model can be extended in various ways using standard techniques. For instance, most of the model parameters were assumed constant over time. It is intuitively reasonable, however, to assume that the behavior of the model parameters is different a hundred days before as compared to a day before the event resolves.  The evolution of the parameters over time could be studied by allowing the parameters to depend on the time index.

Similarly, the model parameters can be allowed to depend on a variety of expert characteristic. In this paper the multiplicative bias was assumed to depend on the level of self-reported expertise. Similar dependencies, however, could be constructed based on any expert characteristic, such as sex, education, specialty, or even a combination of characteristics, to produce a range of novel insight on the level of confidence across different groups of experts.
%The model is also by no means limited to the study of group-level bias at different levels of self-reported expertise. The experts could equally well be partitioned based on other features, such as gender, education, or specialty, to produce a range of novel insight on the level of confidence among different groups of expert forecasters.
Furthermore, instead of grouping the experts, the questions could be grouped by types, such as economic, domestic, or international. These groups could be analyzed separately to learn about forecasting behavior under a wide variety of problem attributes.

%The model could be applied to questions grouped by types, such as economic, domestic, or international, to learn about expert forecasting under a wide variety of problem attributes.

Other future directions could aim to remove some of the obvious limitations of our model. For instance, recall that the random components are assumed to follow a normal distribution. This is a strong assumption that may not always be justified. Logit-probabilities, however, have been modeled with a normal distribution before (see, e.g., \citet{Erev1994}). Furthermore, the normal distribution is a rather standard assumption in psychological models (see, e.g.,  signal-detection theory  in \citet{tanner1954decision}). A second limitation resides in the assumption that both the observed and hidden processes are expected to grow linearly. This assumption could be relaxed, for instance, by adding higher order terms to the model. A more complex model, however, is likely to sacrifice interpretability. Given that our model is able to detect very intricate patterns in the crowd belief  (see Figure \ref{ExamplePlotsFinal}), compromising interpretability for the sake of facilitating non-linear growth is hardly necessary.

>>>>

This paper adds another layer of complexity by assuming a heterogeneous set of experts, most of whom only make one or two probability assessments over the hundred or so days before the event resolves. This means that the decision-maker faces a different group of experts  every day, with only a few experts returning later on for a second round of advice. The problem at hand is therefore strikingly different from many time-series estimation problems, where one has an observation at every time point -- or almost every time point. As a result, standard time-series procedures like ARIMA (see, e.g., \citet{mills1991time}) are not directly applicable to our setting.

In response to the lack of appropriate modeling procedures, this paper introduces a time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated estimate of the crowd belief. The model is highly interpretable and can be used for
\begin{itemize}
\item the analysis of group-level under- and overconfidence across different levels of self-reported expertise,
%\item confidence bands for the crowd belief,
\item accurate probability forecasts, and
\item many question-specific quantities that have easy interpretations and can be used to gain novel insight in the social sciences. 
\end{itemize}

The paper begins with a description of our probability forecasting data that were collected by asking over 2,300 experts to give probability forecasts and to self-assess their level of expertise on a subset of 166 geopolitical binary events. After summarizing the dataset, the paper introduces a dynamic hierarchical model for capturing the crowd belief. The model is estimated in a two-step procedure. First, a sampling step produces constrained parameter estimates via Gibbs sampling (see \citet{geman1984stochastic} for the original introduction of Gibbs sampling). Second, a calibration step transforms these estimates to their unconstrained equivalents via a one-dimension optimization procedure. An extension of this model to polychotomous outcomes is briefly discussed before model evaluation. The first evaluation section uses synthetic data to study the accuracy to which the two-step procedure is able to estimate parameter values. The second evaluation section discusses different aspects of the model by applying it to our real-world probability forecasting data. The paper concludes with a discussion on future research directions and model limitations.

@@

\begin{figure}[h!]
\vspace{-1em}
\centering
\hspace*{0em} 	\includegraphics[width=  \textwidth]{Figures/LegendExamplePlot} % requires the graphicx package
\vspace{-4.5em}

@@

\section{Discussion}
This paper began with an introduction of a rather unorthodox but nonetheless realistic time-series setting in which probability forecasts are made very infrequently by a heterogeneous group of experts. The resulting data is too sparse to be modeled efficiently with standard time-series methods.  In response, our work introduces an interpretable time-series model that incorporates self-reported expertise and captures a sharp and well-calibrated crowd belief across time. The model is estimated in two-steps: The first step, known as the \textit{sampling step}, samples constrained versions of the model parameters via a Gibbs sampler that only makes use of standard distributions. The second step, known as the \textit{calibration step}, involves a one-dimensional optimization procedure that transforms the constrained parameter values to their unconstrained counterparts.

\subsection{Summary of Findings}
The model was applied to an unusually large dataset on probability forecasting. The estimated crowd belief was found to be sharp and well-calibrated under both in- and out-of-sample probability estimation. This has direct implications  on predictive power and model interpretability. Firstly, the model was shown to outperform other probability aggregators in terms of forecasting ability. In particular, the model was deemed a well-balanced compromise that avoids over-fitting without begin overly conservative. Secondly, the crowd belief was used as the no-bias reference point to study the bias among groups of experts with different levels of self-reported expertise. All the groups were found to be under-confident. The under-confidence, however, decreased as the level of self-reported expertise increased. This result is about groups of experts and hence does not conflict with the well-known result of the individuals being over-confident (see, e.g.,  \citet{lichtenstein1977calibration, morgan1992uncertainty, bier2004implications}). Besides making predictions or studying group-level bias, the model can be used to generate estimates of many problem-specific quantities. These quantities have clear interpretations and can be combined in many interesting ways to answer a range of hypotheses about different types of questions and expert behavior.

\subsection{Limitations and Directions for Future Research}
This paper developed a time-series procedure that addresses the main challenges in modeling sparse probability forecasting data. The introduced model was kept parsimonious and can be viewed as a basic template that is amenable to many future extensions. For instance, most of the model parameters were assumed to remain constant over time. It is intuitively reasonable, however, that the behavior of the model parameters is different earlier as opposed to later on in the question.  For instance, the level of disagreement, $\sigma^2_k$, among the experts can be expected to decrease towards the final time point when the the question finally resolves. This hypothesis could be investigated by letting $\sigma^2_{t,k}$ depend on the previous term $\sigma^2_{t-1,k}$ similarly to the hidden process, $X_{t,k}$.

If personal information on the experts were inquired during the data collection process, it is possible to study the forecasting behavior of different kinds of experts. For instance, in this paper the multiplicative bias was modeled as a function of self-reported expertise. The model, however, is by no means restricted to the study of self-reported expertise. Parameter dependencies could be constructed based on many other expert characteristic, such as sex, education or specialty, to produce a range of novel insight on the forecasting behavior across different groups of experts. Furthermore, expert characteristics could be interacted with question types, such as economic, domestic or international, to learn about the kind of experts who generally perform well on certain types of questions. This would be of interest to the decision-maker who could use the information as a basis for consulting only a high-performing subset of the available experts.

Other future directions could aim to remove some of the obvious limitations of our model. For instance, recall that the random components are assumed to follow a normal distribution. This is a strong assumption that may not always be justified. Logit-probabilities, however, have been modeled with the normal distribution before (see, e.g., \citet{Erev1994}). Furthermore, the normal distribution is a rather standard assumption in psychological models (see, e.g.,  signal-detection theory  in \citet{tanner1954decision}). A second limitation resides in the assumption that both the observed and hidden processes are expected to grow linearly. This assumption could be relaxed, for instance, by adding higher order terms to the model. A more complex model, however, is likely to sacrifice interpretability. Given that our model is able to detect very intricate patterns in the crowd belief  (see Figure \ref{ExamplePlotsFinal}), compromising interpretability for the sake of facilitating non-linear growth is hardly necessary.
